---
title: Title
date: 2023-08-23
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true

bibliography: Farkasok.json
---

<!--
  pdf:
    standalone: true
    natbib: true
    biblio-title: References
    fontsize: 12pt
    papersize: a4paper
    header-includes:
      - \usepackage{double_spaced}
      - \usepackage{lineno}
    geometry:
      - margin=1in
    include-before:
    - \linenumbers
-->



```{r load-libs}
library(dplyr)
library(ggplot2)
library(gridExtra)
library(igraph)
library(ggraph)
library(tidygraph)
```


# Grouping by 60 minutes


## Load data

First, load the database.



```{r load-data}
f.df = read.csv2("f_df_en_60.csv")
```

We clean up the data frame, by setting the class of the variables.

```{r clean-df}
fc.df = f.df |>
  mutate(operator = factor(Keszitette),
         location = factor(Helyszin),
         cam.start = as.Date(Start),
         cam.end = as.Date(Stop),
         date = as.Date(Datum),
         rec.start = as.POSIXct(paste(Datum, Felvetel.kezdete),
                                format="%Y-%m-%d %H:%M:%S"),
         rec.end = as.POSIXct(paste(Datum, Felvetel.vege),
                                format="%Y-%m-%d %H:%M:%S"),
         event = factor(Felvetel.tartalma),
         .keep="none"
         ) |>
  filter(cam.start < date & date < cam.end) |>
  arrange(location, cam.start, date) |>
  group_by(location, cam.start) |>
  mutate(time0 = rec.end,
         time1 = lead(rec.start),
         event0 = event,
         event1 = lead(event)
         ) |>
  mutate(time.diff = as.numeric(time1 - time0),
				 time.diff2 = as.numeric(rec.start - lag(rec.end))
         ) |>
  as.data.frame()
ID = paste(fc.df$location, fc.df$cam.start, sep=":")
fc.df$ID = ID
fc.df$recno = 1:nrow(fc.df)
```


Plot the histogram of time differences.

```{r plot-time-diff}
fc.df |>
  ggplot(aes(x = time.diff)) +
  geom_histogram() +
  scale_x_log10()
```


## `msm`

This is based on @jackson2023.

```{r load-msm}
library(msm)
library(minqa)
```

```{r prepare-data-msm}
m.df = data.frame(ID=fc.df$ID,
									recno=fc.df$recno,
									rec.start = fc.df$rec.start,
									time= as.numeric(fc.df$rec.start
																	 - as.POSIXct(as.character(fc.df$cam.start),
																								format="%Y-%m-%d")),
									event.f= fc.df$event0)
d = as.numeric(fc.df$rec.end - fc.df$rec.start)
i = fc.df$time.diff <= 0 | d < 0 | d > 3600
m.df = m.df[!i,]
m.df$event.f[m.df$event.f == "rider"] = "hiker"
m.df = droplevels(m.df)
m.df$event = as.numeric(m.df$event.f)
m.df = m.df[order(m.df$ID, m.df$time),]
m.df = m.df[!(m.df$ID == "Harsas02:2018-06-20" | m.df$ID ==
            "Sugerkut:2015-10-07"),]
```

```{r state-matrix}
statetable.msm(event.f, ID, m.df)
statetable.msm(event, ID, m.df)
```

```{r Q-matrix}
Q = matrix(rep(1/7, 49), ncol=7)
```

```{r crude-estimates}
Q.crude = crudeinits.msm(event ~ time, ID, data=m.df, qmatrix=Q)
```

First fit.

```{r first-msm-fit}
fname = "msm_transition_intesities-60min.Rdata"
if(file.exists(fname)) {
	load(fname)
} else {
	m0 = msm(event ~ time, subject=ID, data=m.df, qmatrix=Q.crude,
					 opt.method="bobyqa", control=list(maxfun=1e6))
	save(m0, file=fname)
}
```

### Plotting transition probabilities

```{r prepare-data-transition}
times = 10^seq(-2,1, 0.25)
g = expand.grid(from=levels(m.df$event.f), to=levels(m.df$event.f))
r = data.frame()
for(tt in times) {
	pm = pmatrix.msm(m0, tt)
	d = data.frame(from=g$from, to=g$to, time=tt, prob=as.numeric(pm))
	r = rbind(r, d)
}
```

```{r plot-transition-probs2}
r |> filter(time <= 1) |>
ggplot(aes(x=time, y=prob, group=from, colour=from)) +
	geom_point(aes(shape=from)) +
	geom_line() +
	scale_x_log10() +
	scale_y_log10() +
	#theme(axis.text.x=element_text(angle=45, hjust=1)) +
	facet_wrap(~to, labeller="label_both")
```

```{r plot-transition-probs-column}
r |> filter(time == 0.01) |>
  ggplot(aes(x=from, y=prob, group=to, fill=from)) +
    geom_line() +
    geom_point() +
    scale_y_log10() +
    theme(axis.text.x=element_text(angle=45, hjust=1)) +
    facet_wrap(~to, labeller="label_both")
```

### Plotting the probabilities of next states

```{r calc-next-state-probs}
pnext = pnext.msm(m0)
pnext.df = data.frame(current.st=g$from, next.st=g$to,
                      prob.next=as.numeric(pnext$estimates),
                      prob.next.L=as.numeric(pnext$L),
                      prob.next.U=as.numeric(pnext$U))
pnext.df$current.type = "ungulates"
pnext.df$current.type[grepl("(Vulpes|Canis)", pnext.df$current.st)] = "predator"
pnext.df$current.type[grepl("(hiker|MV)", pnext.df$current.st)] = "human"
pnext.df$current.type = factor(pnext.df$current.type)
pnext.df$next.type = "ungulates"
pnext.df$next.type[grepl("(Vulpes|Canis)", pnext.df$next.st)] = "predator"
pnext.df$next.type[grepl("(hiker|MV)", pnext.df$next.st)] = "human"
pnext.df$next.type = factor(pnext.df$next.type)
pnext.df$current.st = factor(pnext.df$current.st,
                             levels=c("hiker", "MV", "Canis lupus",
                                      "Vulpes vulpes", "Capreolus capreolus",
                                      "Cervus elaphus", "Sus scrofa"))
pnext.df$next.st = factor(pnext.df$next.st,
                             levels=c("hiker", "MV", "Canis lupus",
                                      "Vulpes vulpes", "Capreolus capreolus",
                                      "Cervus elaphus", "Sus scrofa"))
pnext.df$prob.next.L[pnext.df$prob.next <= 0] = NA
pnext.df$prob.next.U[pnext.df$prob.next <= 0] = NA
pnext.df$prob.next[pnext.df$prob.next <= 0] = NA
sojourn = sojourn.msm(m0)
sojourn$state = levels(m.df$event.f)
sojourn$state = factor(sojourn$state,
                             levels=c("hiker", "MV", "Canis lupus",
                                      "Vulpes vulpes", "Capreolus capreolus",
                                      "Cervus elaphus", "Sus scrofa"))
sojourn$type = c("predator", "ungulate", "ungulate", "human", "human",
                 "ungulate", "predator")
```

```{r matplot-next-state-probs}
pnext.df$prob.next[is.na(pnext.df$prob.next)] = 0
pnext.gg = ggplot(pnext.df, aes(x=next.st, y=current.st, fill=prob.next)) +
  geom_tile() +
  scale_fill_gradient(low="white", high="red") +
  geom_text(aes(x=next.st, y=current.st, label=round(prob.next, 3))) +
  scale_y_discrete(limits=rev) +
  scale_x_discrete(position="top") +
	theme(axis.text.x=element_text(angle=45, hjust=0)) +
  labs(title="(b) probability of next state", x="next state",
       y="current state", fill="probability\nof\nnext\nstate") +
  theme(aspect.ratio=1, plot.title=element_text(hjust=0.5))
pnext.gg

sojourn.gg = ggplot(sojourn, aes(x=state, y=estimates, fill=type)) +
  geom_bar(stat="identity") +
  geom_errorbar(aes(ymin=L, ymax=U), width=0.2) +
  theme(axis.text.x=element_text(angle=45, hjust=1),
        plot.title=element_text(hjust=0.5)) +
  labs(title="(a) sojourn times", y="sojourn time (days)")

sojourn.gg

##grid.arrange(sojourn.gg, pnext.gg, layout_matrix=matrix(1:2, ncol=2))
```


### Network on the basis of next state probabilities and sojourns

```{r create-graph-sojourn}
g.qr = graph_from_adjacency_matrix(pnext$estimate, mode="directed",
																	 weighted=TRUE, diag=FALSE)
V(g.qr)$name = levels(m.df$event.f)
V(g.qr)$size = sojourn$estimate
V(g.qr)$type = c("predator", "ungulate", "ungulate", "human", "human",
                 "ungulate", "predator")
```

```{r plot-graph-sojourn, eval=FALSE}
plot(g.qr, edge.width=10*E(g.qr)$weight, edge.curved=TRUE,
     edge.arrow.size=1.5, vertex.color=c(1, 2, 2, 3, 3, 2, 1),
     vertex.size=V(g.qr)$size*250)
```

```{r ggplot-graph-sojourn}

set.seed(4)
layout = create_layout(g.qr, layout="nicely")
set.seed(NULL)
##graph.gg = ggraph(g.qr, layout="nicely") +
graph.gg = ggraph(layout) +
  ggtitle("(c) transition diagram") +
  geom_edge_fan(aes(width=weight), colour="grey", alpha=0.5,
                #strength=0.95,
                arrow=arrow(length=unit(4, "mm"), type="closed"),
                #start_cap=square(3, 'mm'),
                end_cap=circle(10, "mm"), show.legend=FALSE) +
  geom_edge_fan(aes(width=weight), #strength=0.95,
                alpha=0.5,
                show.legend=FALSE,
                colour="grey") +
  scale_edge_width(range=c(0.01, 3)) +
  #geom_node_point(size=250*V(g.qr)$size, aes(colour=type)) +
  geom_node_point(aes(size=size, colour=type), show.legend=FALSE) +
  scale_radius(range=c(10,25)) +
  geom_node_text(aes(label=name)) +
  theme_void() +
  theme(aspect.ratio=1, plot.title=element_text(hjust=0.5))
graph.gg

pdf(file="transitions-60min.pdf", width=15, height=12)
grid.arrange(sojourn.gg, pnext.gg, graph.gg,
             layout_matrix=matrix(c(1, 2, 3, 3), ncol=2))
dev.off()

```

# Grouping by 15 minutes


## Load data

First, load the database.



```{r load-data}
f.df = read.csv("bnpi_wolf_camtrap/Újragondolt/f_df_en.csv")
```

We clean up the data frame, by setting the class of the variables.

```{r clean-df}
fc.df = f.df |>
  mutate(operator = factor(Keszitette),
         location = factor(Helyszin),
         cam.start = as.Date(Start),
         cam.end = as.Date(Stop),
         date = as.Date(Datum),
         rec.start = as.POSIXct(paste(Datum, Felvetel.kezdete),
                                format="%Y-%m-%d %H:%M:%S"),
         rec.end = as.POSIXct(paste(Datum, Felvetel.vege),
                                format="%Y-%m-%d %H:%M:%S"),
         event = factor(Felvetel.tartalma),
         .keep="none"
         ) |>
  filter(cam.start < date & date < cam.end) |>
  arrange(location, cam.start, date) |>
  group_by(location, cam.start) |>
  mutate(time0 = rec.end,
         time1 = lead(rec.start),
         event0 = event,
         event1 = lead(event)
         ) |>
  mutate(time.diff = as.numeric(time1 - time0),
				 time.diff2 = as.numeric(rec.start - lag(rec.end))
         ) |>
  as.data.frame()
ID = paste(fc.df$location, fc.df$cam.start, sep=":")
fc.df$ID = ID
fc.df$recno = 1:nrow(fc.df)
```


Plot the histogram of time differences.

```{r plot-time-diff}
fc.df |>
  ggplot(aes(x = time.diff)) +
  geom_histogram() +
  scale_x_log10()
```


## `msm`

This is based on @jackson2023.

```{r load-msm}
library(msm)
library(minqa)
```

```{r prepare-data-msm}
m.df = data.frame(ID=fc.df$ID,
									recno=fc.df$recno,
									rec.start = fc.df$rec.start,
									time= as.numeric(fc.df$rec.start
																	 - as.POSIXct(as.character(fc.df$cam.start),
																								format="%Y-%m-%d")),
									event.f= fc.df$event0)
d = as.numeric(fc.df$rec.end - fc.df$rec.start)
i = fc.df$time.diff <= 0 | d < 0 | d > 3600
m.df = m.df[!i,]
m.df$event.f[m.df$event.f == "rider"] = "hiker"
m.df = droplevels(m.df)
m.df$event = as.numeric(m.df$event.f)
m.df = m.df[order(m.df$ID, m.df$time),]
m.df = m.df[!(m.df$ID == "Harsas02:2018-06-20" | m.df$ID ==
            "Sugerkut:2015-10-07"),]
```

```{r state-matrix}
statetable.msm(event.f, ID, m.df)
statetable.msm(event, ID, m.df)
```

```{r Q-matrix}
Q = matrix(rep(1/7, 49), ncol=7)
```

```{r crude-estimates}
Q.crude = crudeinits.msm(event ~ time, ID, data=m.df, qmatrix=Q)
```

First fit.

```{r first-msm-fit}
fname = "msm_transition_intesities.Rdata"
if(file.exists(fname)) {
	load(fname)
} else {
	m0 = msm(event ~ time, subject=ID, data=m.df, qmatrix=Q.crude,
					 opt.method="bobyqa", control=list(maxfun=1e6))
	save(m0, file=fname)
}
```

### Plotting transition probabilities

```{r prepare-data-transition}
times = 10^seq(-2,1, 0.25)
g = expand.grid(from=levels(m.df$event.f), to=levels(m.df$event.f))
r = data.frame()
for(tt in times) {
	pm = pmatrix.msm(m0, tt)
	d = data.frame(from=g$from, to=g$to, time=tt, prob=as.numeric(pm))
	r = rbind(r, d)
}
```

```{r plot-transition-probs2}
r |> filter(time <= 1) |>
ggplot(aes(x=time, y=prob, group=from, colour=from)) +
	geom_point(aes(shape=from)) +
	geom_line() +
	scale_x_log10() +
	scale_y_log10() +
	#theme(axis.text.x=element_text(angle=45, hjust=1)) +
	facet_wrap(~to, labeller="label_both")
```

```{r plot-transition-probs-column}
r |> filter(time == 0.01) |>
  ggplot(aes(x=from, y=prob, group=to, fill=from)) +
    geom_line() +
    geom_point() +
    scale_y_log10() +
    theme(axis.text.x=element_text(angle=45, hjust=1)) +
    facet_wrap(~to, labeller="label_both")
```

### Plotting the probabilities of next states

```{r calc-next-state-probs}
pnext = pnext.msm(m0)
pnext.df = data.frame(current.st=g$from, next.st=g$to,
                      prob.next=as.numeric(pnext$estimates),
                      prob.next.L=as.numeric(pnext$L),
                      prob.next.U=as.numeric(pnext$U))
pnext.df$current.type = "ungulates"
pnext.df$current.type[grepl("(Vulpes|Canis)", pnext.df$current.st)] = "predator"
pnext.df$current.type[grepl("(hiker|MV)", pnext.df$current.st)] = "human"
pnext.df$current.type = factor(pnext.df$current.type)
pnext.df$next.type = "ungulates"
pnext.df$next.type[grepl("(Vulpes|Canis)", pnext.df$next.st)] = "predator"
pnext.df$next.type[grepl("(hiker|MV)", pnext.df$next.st)] = "human"
pnext.df$next.type = factor(pnext.df$next.type)
pnext.df$current.st = factor(pnext.df$current.st,
                             levels=c("hiker", "MV", "Canis lupus",
                                      "Vulpes vulpes", "Capreolus capreolus",
                                      "Cervus elaphus", "Sus scrofa"))
pnext.df$next.st = factor(pnext.df$next.st,
                             levels=c("hiker", "MV", "Canis lupus",
                                      "Vulpes vulpes", "Capreolus capreolus",
                                      "Cervus elaphus", "Sus scrofa"))
pnext.df$prob.next.L[pnext.df$prob.next <= 0] = NA
pnext.df$prob.next.U[pnext.df$prob.next <= 0] = NA
pnext.df$prob.next[pnext.df$prob.next <= 0] = NA
sojourn = sojourn.msm(m0)
sojourn$state = levels(m.df$event.f)
sojourn$state = factor(sojourn$state,
                             levels=c("hiker", "MV", "Canis lupus",
                                      "Vulpes vulpes", "Capreolus capreolus",
                                      "Cervus elaphus", "Sus scrofa"))
sojourn$type = c("predator", "ungulate", "ungulate", "human", "human",
                 "ungulate", "predator")
```

```{r matplot-next-state-probs}
pnext.df$prob.next[is.na(pnext.df$prob.next)] = 0
pnext.gg = ggplot(pnext.df, aes(x=next.st, y=current.st, fill=prob.next)) +
  geom_tile() +
  scale_fill_gradient(low="white", high="red") +
  geom_text(aes(x=next.st, y=current.st, label=round(prob.next, 2))) +
  scale_y_discrete(limits=rev) +
  scale_x_discrete(position="top") +
	theme(axis.text.x=element_text(angle=45, hjust=0)) +
  labs(title="(b) probability of next state", x="next state",
       y="current state", fill="probability\nof\nnext\nstate") +
  theme(aspect.ratio=1, plot.title=element_text(hjust=0.5))
##pnext.gg
sojourn.gg = ggplot(sojourn, aes(x=state, y=estimates, fill=type)) +
  geom_bar(stat="identity") +
  geom_errorbar(aes(ymin=L, ymax=U), width=0.2) +
  theme(axis.text.x=element_text(angle=45, hjust=1),
        plot.title=element_text(hjust=0.5)) +
  labs(title="(a) sojourn times", y="sojourn time (days)")
##sojourn.gg

##grid.arrange(sojourn.gg, pnext.gg, layout_matrix=matrix(1:2, ncol=2))
```


### Network on the basis of next state probabilities and sojourns

```{r create-graph-sojourn}
g.qr = graph_from_adjacency_matrix(pnext$estimate, mode="directed",
																	 weighted=TRUE, diag=FALSE)
V(g.qr)$name = levels(m.df$event.f)
V(g.qr)$size = sojourn$estimate
V(g.qr)$type = c("predator", "ungulate", "ungulate", "human", "human",
                 "ungulate", "predator")
```

```{r plot-graph-sojourn, eval=FALSE}
plot(g.qr, edge.width=10*E(g.qr)$weight, edge.curved=TRUE,
     edge.arrow.size=1.5, vertex.color=c(1, 2, 2, 3, 3, 2, 1),
     vertex.size=V(g.qr)$size*250)
```

```{r ggplot-graph-sojourn}
set.seed(8)
layout = create_layout(g.qr, layout="nicely")
set.seed(NULL)
##graph.gg = ggraph(g.qr, layout="nicely") +
graph.gg = ggraph(layout) +
  ggtitle("(c) transition diagram") +
  geom_edge_fan(aes(width=weight), colour="grey", #alpha=0.25,
                #strength=0.95,
                arrow=arrow(length=unit(4, "mm"), type="closed"),
                #start_cap=square(3, 'mm'),
                end_cap=circle(10, "mm"), show.legend=FALSE) +
  geom_edge_fan(aes(width=weight), #strength=0.95,
                #alpha=0.5,
                show.legend=FALSE,
                colour="grey") +
  scale_edge_width(range=c(0.1, 2)) +
  #geom_node_point(size=250*V(g.qr)$size, aes(colour=type)) +
  geom_node_point(aes(size=size, colour=type), show.legend=FALSE) +
  scale_radius(range=c(10,25)) +
  geom_node_text(aes(label=name)) +
  theme_void() +
  theme(aspect.ratio=1, plot.title=element_text(hjust=0.5))
graph.gg

pdf(file="transitions.pdf", width=15, height=12)
grid.arrange(sojourn.gg, pnext.gg, graph.gg,
             layout_matrix=matrix(c(1, 2, 3, 3), ncol=2))
dev.off()

```

# End matters

```{r session-info, include=TRUE, echo=TRUE, results="markup"}
sessionInfo()
```

<!-- vim: set foldmethod=syntax: -->
